\documentclass[]{article}

\usepackage{times,amsmath}

% revise margins
\setlength{\headheight}{0.0in}
\setlength{\topmargin}{0.0in}
\setlength{\headsep}{0.0in}
\setlength{\textheight}{9.0in}
\setlength{\footskip}{0.0in}
\setlength{\oddsidemargin}{0.0in}
\setlength{\evensidemargin}{0.0in}
\setlength{\textwidth}{6.5in}
\setlength{\parindent}{0in}

\newcommand{\hsp}{\hspace*{0.2cm}}

\pagestyle{empty}

\begin{document}


{ \large
Assignment 1 \hfill 140.778 (Adv Stat Comp)
\\
Due: 29 Nov \hfill 2nd term, 2000--2001 }

\bigskip \bigskip

{\large \textbf{The bootstrap in linear regression}}

\bigskip 

\textbf{Part A}

\smallskip

Consider the model $y_i = \beta_0 + \beta_1 x_i + \varepsilon_i$, $i = 1,
\ldots, n$, where 
\begin{itemize}
  \item $x_i \sim \text{ iid U}(0,10)$ \hsp or \\
      $x_i \sim \text{ iid N}(\mu=5, \sigma^2=3)$
  \item $\varepsilon_i \sim \text{ iid N}(0, \sigma^2)$ \hsp or \\
        $\varepsilon_i \sim \text{ indep N}(0, \sigma^2x_i)$ \hsp or \\
        $\varepsilon_i/\gamma \sim \text{ iid t}(\text{df} = 3)$ \hsp or \\
        $\varepsilon_i + 1/\lambda \sim \text{ iid exponential}(\lambda)$
\end{itemize}

Write a function (in R, S-plus, or your favorite such package) to
simulate such data, where you allow the user to specify $n$, the
models for the $x_i$ and the $\varepsilon_i$ and the unknown
parameters $\beta_0$, $\beta_1$ and $\sigma^2$, $\gamma$ or
$\lambda$.  

\bigskip

\textbf{Part B}

\smallskip

\begin{enumerate}

\item Download the comma-delimited file \verb|data1.csv| from the
  course web page. 

\item Fit the linear model $y = \beta_0 + \beta_1 x + \varepsilon$
  where the $\varepsilon$ are assumed to be iid N$(0, \sigma^2)$.
  Obtain, using the normal theory, estimates and estimated standard
  errors (SEs) of $\hat{\beta}_0$ and $\hat{\beta}_1$ and confidence
  intervals (CIs) for $\beta_0$ and $\beta_1$.  Also get an estimate
  of the correlation between $\hat{\beta}_0$ and $\hat{\beta}_1$.  

\item Use a parametric bootstrap (ie, Monte Carlo simulation), using
  the normal model, to get new SEs and CIs and estimated correlation.

\item Use a nonparametric bootstrap, resampling the residuals, to get
  another set of SEs, etc.

\item Use a nonparametric bootstrap, resampling the pairs $(x,y)$, to
  get a final set of SEs, etc.

\item In at least one of 3--5, try to estimate the Monte Carlo error
  in the estimated SEs and CIs.

\item Discuss the results.  

\end{enumerate}

\bigskip

\textbf{Part C}

Use your simulation program from A to investigate the performance of
the parametric bootstrap and the two nonparametric bootstraps as a
function of the sample size $n$, the model for $\varepsilon_i$ and
perhaps the model for the $x_i$.

\bigskip

\textbf{Notes}: 
\begin{enumerate}
\item You do not need to complete all aspects of this assignment in
  great detail.  Use your judgement in deciding how much time you wish
  to devote to it (considering your other responsibilities, your
  interest, and your need for sleep).
\item In reporting your results, \emph{do not\/} simply provide raw
  computer output.  I expect to see the sort of report one might give
  to a consulting client, with any computer code that you feel you
  should include placed in an appendix.
\item Feel free to ask me for advice, and feel free to work together.
  The report, however, should be your own.
\end{enumerate}

\bigskip

\textbf{Useful R functions}: \\
\verb|summary(a <- lm(y~x))|, \verb|sample|, \verb|rnorm|,
\verb|rexp|, \verb|rt|, \verb|mean|, \verb|sd|, \verb|cor|, \verb|for|

\end{document}
