\documentclass[12pt]{article}

\usepackage{times,amsmath}

% revise margins
\setlength{\headheight}{0.0in}
\setlength{\topmargin}{0.0in}
\setlength{\headsep}{0.0in}
\setlength{\textheight}{9.0in}
\setlength{\footskip}{0.0in}
\setlength{\oddsidemargin}{0.0in}
\setlength{\evensidemargin}{0.0in}
\setlength{\textwidth}{6.5in}
\setlength{\parindent}{0in}

\newcommand{\hsp}{\hspace*{0.2cm}}

\pagestyle{empty}

\begin{document}


{ \large
Assignment 2 \hfill 140.778 (Adv Stat Comp)
\\
Due: 20 Dec \hfill 2nd term, 2000--2001 }

\bigskip \bigskip

{\large \textbf{EM algorithm}}

\bigskip 

\textbf{Part A}

\bigskip

Consider the linear model $y^\star = X \beta + \varepsilon$ where
$\varepsilon \sim \text{N}(0, \sigma^2I)$.  

\bigskip

Suppose that $X$ is observed, but that rather than observing $y_i^\star$, we
observe $y_i = \min\{ y_i^\star, c \}$ (where the censoring value $c$
is known and is constant for all $i$).

\bigskip

Work out an EM algorithm to obtain the MLEs of $\beta$ and $\sigma$.
(You will no doubt find it very useful to look at problem 10.8 in K
Lange, \emph{Numerical analysis for statisticians}, pp 126--127.)  

\bigskip

Write an R (or Splus, or even Matlab) function to implement your
algorithm.  

\bigskip \bigskip



\textbf{Part B}

\smallskip

\begin{enumerate}

\item Download the comma-delimited file \verb|data2.csv| from the
  course web page.  The $y$'s are censored at 10.0. 

\item Fit the linear model $y = X\beta + \varepsilon$, ignoring the
  censoring.

\item Use your EM function to get the MLEs of $\beta$ and $\sigma$,
  taking account of the censoring.

\item Consider playing with starting values.  Does the likelihood
  surface have multiple modes?

\item If time permits, try using a parametric and/or nonparametric
  bootstrap to get SEs of $\beta$ and $\sigma$.  

\item If time permits, write a function to calculate the observed-data
  log likelihood, and apply the \verb|nlm| (or \verb|nlmin|) function
  to get MLEs.  Using \verb|hessian=TRUE| will allow you to get an
  estimated variance matrix.

\item Consider comparing the number of iterations, computer time, and
  sensitivity to starting values for the EM vs \verb|nlm| approaches.

\end{enumerate}


\bigskip \bigskip

\textbf{Useful R functions}: \\
\verb|read.table|, \verb|nlm|, \verb|lm|,
\verb|dnorm|, \verb|pnorm|, \verb|unix.time|

\end{document}
